---
title: 'Lab 1: Species Distribution Modeling - Explore'
author: "Clarissa Boyajian"
date: "1/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load packages, installing if missing
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}

librarian::shelf(
  tidyverse, dismo, DT, here, htmltools, leaflet, mapview, purrr, raster, 
  readr, rgbif, rgdal, rJava, sdmpredictors, sf, spocc, tidyr, geojsonio)
select <- dplyr::select # overwrite raster::select
options(readr.show_col_type = FALSE)

# set random seed for reproducibility
set.seed(42)

# directory to store data
dir_data <- here("data/sdm")
dir.create(dir_data, showWarnings = F, recursive = TRUE)
```


## Get Species Observations

```{r}
obs_csv <- file.path(dir_data, "obs.csv")
obs_geo <- file.path(dir_data, "obs.geojson")
redo    <- TRUE

if (!file.exists(obs_geo) | redo){
  # get species occurrence data from GBIF with coordinates
  (res <- spocc::occ(
    query = 'Buteo jamaicensis',
    from = 'gbif',
    has_coords = T,
    limit = 10000))
  
  # extract data frame from result
  df <- res$gbif$data[[1]] 
  readr::write_csv(df, obs_csv)

  # convert to points of observation from lon/lat columns in data frame
  obs <- df %>% 
    # limit observations to bounding box around north and central america
    filter(between(longitude, -167.593385, -51.171266),
           between(latitude, 5.645215, 71.374349)) %>% 
    sf::st_as_sf(
      coords = c("longitude", "latitude"),
      crs = st_crs(4326)) %>% 
    select(prov, key) %>%  # save space (joinable from obs_csv)
    distinct(geometry, .keep_all = TRUE) # remove duplicate geometries
  
  sf::write_sf(obs, obs_geo, delete_dsn = T)
}

obs <- sf::read_sf(obs_geo)
nrow(obs) # number of rows
```

```{r}
# show points on map
mapview::mapview(obs, map.types = "Esri.WorldStreetMap")
```


#### Questions:
**Question 1**: There were a total of 7,004,451 observations in GBIF for the Red-tailed hawk (Buteo jamaicensis) as of 2022-01-24.
**Question 2**: There was one point from the original dataset that was in the ocean near Europe. To fix this issues, I created a bounding box around the parts of North and Central America and used this bbox to exclude any locations outside of the Red-tailed Hawks habitat.


## Get Environmental Data

### Presence

```{r}
dir_env <- file.path(dir_data, "env")

# set a default data directory
options(sdmpredictors_datadir = dir_env)

# choosing terrestrial
env_datasets <- sdmpredictors::list_datasets(terrestrial = TRUE, marine = FALSE)

# show table of datasets
env_datasets %>% 
  select(dataset_code, description, citation) %>% 
  DT::datatable()
```

```{r}
# choose datasets for a vector
env_datasets_vec <- c("WorldClim", "ENVIREM")

# get layers
env_layers <- sdmpredictors::list_layers(env_datasets_vec)
DT::datatable(env_layers)
```

```{r}
# choose layers after some inspection and perhaps consulting literature
env_layers_vec <- c("WC_alt", # altitude
                    "WC_bio1", # annual mean temperature
                    "WC_bio2", # mean diurnal temperature range
                    "ER_tri", # terrain roughness
                    "WC_bio12") # annual precipitation

# get layers
env_stack <- load_layers(env_layers_vec)

# interactive plot layers, hiding all but first (select others)
# mapview(env_stack, hide = T) # makes the html too big for Github
plot(env_stack, nc = 2)
```

```{r}
obs_hull_geo  <- file.path(dir_data, "obs_hull.geojson")
env_stack_grd <- file.path(dir_data, "env_stack.grd")

if (!file.exists(obs_hull_geo) | redo){
  # make convex hull around points of observation
  obs_hull <- sf::st_convex_hull(st_union(obs))
  
  # save obs hull
  write_sf(obs_hull, obs_hull_geo)
}
obs_hull <- read_sf(obs_hull_geo)

# show points on map
mapview(
  list(obs, obs_hull))
```

```{r}
if (!file.exists(env_stack_grd) | redo){
  obs_hull_sp <- sf::as_Spatial(obs_hull)
  env_stack <- raster::mask(env_stack, obs_hull_sp) %>% 
    raster::crop(extent(obs_hull_sp))
  writeRaster(env_stack, env_stack_grd, overwrite = T)  
}
env_stack <- stack(env_stack_grd)

# show map
plot(env_stack, nc=2)
```

### Pseudo-Absence

```{r}
absence_geo <- file.path(dir_data, "absence.geojson")
pts_geo     <- file.path(dir_data, "pts.geojson")
pts_env_csv <- file.path(dir_data, "pts_env.csv")

if (!file.exists(absence_geo) | redo){
  # get raster count of observations
  r_obs <- rasterize(
    sf::as_Spatial(obs), env_stack[[1]], field=1, fun='count')
  
  # create mask for 
  r_mask <- mask(env_stack[[1]] > -Inf, r_obs, inverse=T)
  
  # generate random points inside mask
  absence <- dismo::randomPoints(r_mask, nrow(obs)) %>% 
    as_tibble() %>% 
    st_as_sf(coords = c("x", "y"), crs = 4326)
  
  write_sf(absence, absence_geo, delete_dsn = T)
}
absence <- read_sf(absence_geo)

# show map of presence, ie obs, and absence
mapview(obs, col.regions = "green") + 
  mapview(absence, col.regions = "gray")
```

```{r}
if (!file.exists(pts_env_csv) | redo){

  # combine presence and absence into single set of labeled points 
  pts <- rbind(
    obs %>% 
      mutate(present = 1) %>% 
      select(present, key),
    absence %>% mutate(present = 0, 
                       key = NA)) %>% 
    mutate(ID = 1:n()) %>% 
    relocate(ID)
  write_sf(pts, pts_geo, delete_dsn = TRUE)

  # extract raster values for points
  pts_env <- raster::extract(env_stack, as_Spatial(pts), df = TRUE) %>% 
    tibble() %>% 
    # join present and geometry columns to raster value results for points
    left_join(pts %>% select(ID, present), 
              by = "ID") %>% 
    relocate(present, .after = ID) %>% 
    # extract lon, lat as single columns
    mutate(lon = st_coordinates(geometry)[,1],
           lat = st_coordinates(geometry)[,2]) %>% 
    select(-geometry)
  write_csv(pts_env, pts_env_csv)
}
pts_env <- read_csv(pts_env_csv)

pts_env %>% 
  # show first 10 presence, last 10 absence
  slice(c(1:10, (nrow(pts_env)-9):nrow(pts_env))) %>% 
  DT::datatable(rownames = FALSE,
                options = list(dom = "t",
                               pageLength = 20))
```


## Term Plots

```{r}
pts_env %>% 
  select(-ID) %>% 
  mutate(present = factor(present)) %>% 
  pivot_longer(-present) %>% 
  ggplot() +
  geom_density(aes(x = value, fill = present)) + 
  scale_fill_manual(values = alpha(c("gray", "green"), 0.5)) +
  scale_x_continuous(expand=c(0,0)) +
  scale_y_continuous(expand=c(0,0)) +
  theme_bw() + 
  facet_wrap(~name, scales = "free") +
  theme(legend.position = c(1, 0),
        legend.justification = c(1, 0))
```




